---
title: "Linear Regression"
format:
  revealjs:
    scrollable: true
    navigation-mode: vertical
    controls-layout: bottom-right
    controls-tutorial: true
    incremental: true 
    chalkboard:
      src: chalkboard.json
      storage: "chalkboard_pres"
      theme: whiteboard
      chalk-width: 4
editor: source
---

```{r}
#| include: false
library(tidyverse)
penguins <- penguins |> drop_na()

```

## Learning Outcomes

- Scatter Plot

- Linear Regression

- Ordinary Least Squares

- Unbiasedness

# Scatter Plot

## Scatter Plot

```{r}
ggplot(penguins, aes(y = flipper_len, x = body_mass)) +
  geom_point() + theme_bw()
```

## Scatter Plot

```{r}
sample_n(penguins,10) |> 
  ggplot(aes(y = flipper_len, x = body_mass)) +
    geom_point() + 
    theme_bw()
```



# Linear Regression

## Linear Regression

Linear regression is used to model the association between a set of predictor variables (x's) and an outcome variable (y). Linear regression will fit a line that best describes the data points.

## Simple Linear Regression

Simple linear regression will model the association between one predictor variable and an outcome:

$$
Y = \beta_0 + \beta_1 X + \epsilon
$$

- $\beta_0$: Intercept term

- $\beta_1$: Slope term

- $\epsilon\sim N(0,\sigma^2)$ 


## Fitting a Line 

```{r}
penguins |> 
  ggplot(aes(y = flipper_len, x = body_mass)) + 
    geom_point() + 
    geom_smooth(method = "lm") +
    theme_bw() + 
    annotate("text", label = "y=136.7+0.015x", x=3250, y=230, size = 10)
```
## Interpretation

$$
\hat y = 136.73 + 0.015 x
$$

# Ordinary Least Squares

## Ordinary Least Squares

For a data pair $(X_i,Y_i)_{i=1}^n$, the ordinary least squares estimator will find the estimates of $\hat\beta_0$ and $\hat\beta_1$ that minimize the following function:

$$
\sum^n_{i=1}\{y_i-(\beta_0+\beta_1x_i)\}^2
$$

## Estimating $\beta$'s

## Estimating $\beta_1$

## Estimating $\beta_0$

## Estimates

$$
\hat\beta_0 = \bar y - \hat\beta_1\bar x
$$
$$
\hat\beta_1 = \frac{\sum^n_{i=1}(y_i-\bar y)(x_i-\bar x)}{\sum^n_{i=1}(x_i-\bar x)^2}
$$
$$
\hat\sigma^2 = \frac{1}{n-2}\sum^n_{i=1}(y_i-\hat y_i)^2
$$

# Unbiasedness of $\beta$'s

## Unbiasedness of $\beta$'s

Both $\beta_0$ and $\beta_1$ are unbiased estimators.

##  $E(\beta_1)$

##  $E(\beta_0)$
